---
title: "Virtual Reality Interventions: Evaluating Efficacy in the Treatment of GAD"
author: "Alan Kinsela, Erik Trujillo"
date: "13/02/2023"
output: html_document
---

# Setup 
```{r include=FALSE}

# install.packages("tidyverse")
if(!require("knitr"))
library(knitr)

# Load the packages, loading tidyverse for readr, dplyr, magrittr, and ggplot2
if(!require("tidyverse"))
library(tidyverse)
```


```{r} 
#Setting this up to define all chunks to show both code and output: 
knitr::opts_chunk$set(echo = TRUE)

```

# Data Loading
Loading our Datasets into R Studio
```{r} 
control_group <- read_csv("GROUP_12_2023_GCA_RESULTS_CONTROL.csv")
experimental_group <- read_csv("GROUP_12_2023_GCA_RESULTS_EXPERIMENTAL.csv")

```

# Initial Inspection
Initial inspection shows the same data types on both datasets. Patients are categorized by gender, and there are numeric scores for Generalized Anxiety Disorder (GAD) and State-Trait Anxiety Inventory (STAI). We have results both for pretrial testing and post trial testing. 
```{r} 
# Display the first few rows of the datasets
 head(control_group)
 head(experimental_group)
```

The summary is showing some insight on how the data is distributed, there are also some outliers within minimum values in scores that might need to be looked at during our cleaning process. 
```{r} 
# Summary of the datasets 
 summary(control_group)
 summary(experimental_group)
```

# Data Cleanup
```{r include=FALSE}
# Check for missing values
cntlMissingCount <- sum(is.na(control_group)) 
expMissingCount <- sum(is.na(experimental_group)) 

# Verifying data types 
cntlDataTypes <- str(control_group) 
expDataTypes <- str(experimental_group) 

```

During the data cleanup stage we check for any missing values and found `r cntlMissingCount` missing values in the control group and `r expMissingCount` missing values in the experimental group meaning no participants need to be thrown out for lack of data.

Upon inspecting data types by viewing the structure of the data, all data types match what they should:

- Patient ID, and test results are all numbers
- Gender are all character strings

<!-- This is what a comment looks like-->
### Table of Contents

1. [Abstract](#abstract)
2. [Introduction](#introduction)
3. [Method](#method)
4. [Results](#results)
5. [Discussion](#discussion)
6. [References](#references)


### Abstract <a name="abstract"></a>



### Introduction <a name="introduction"></a>




### Method <a name="method"></a>




### Results <a name="results"></a>




### Discussion <a name="discussion"></a>




### References <a name="references"></a>


Code for Knitr: https://argoshare.is.ed.ac.uk/healthyr_book/the-anatomy-of-a-notebook-r-markdown-file.html